### 2025-07-02 machine-learning.qmd
20250702 - machine learning

```{r}
library("petersenlab")
library("parallel")
library("doParallel")
library("missRanger")
library("powerjoin")
library("tidymodels")
library("LongituRF")
#library("caret")
library("gpboost")
library("tidyverse")
```

@@

num_cores <- detectCores() - 1
num_true_cores <- parallel::detectCores(logical = FALSE) - 1
```

@@

@@

@@

@@

@@

@@

@@

We use the `caret` package [@R-caret].
We use the `doParallel` [@R-doParallel] package for parallel (faster) processing.

```{r}
#| label: cross-sectional-machine-learning
#| eval: false

cl <- parallel::makeCluster(num_cores)
doParallel::registerDoParallel(cl)

set.seed(52242)

randomForest_qb <- caret::train(
  fantasyPoints_lag ~ ., # use all predictors
  data = seasonalData_lag_qb_train_imp$data,
  method = "rf",
  trControl = trainControl(
    method = "cv",
    number = 10)) # 10-fold cross-validation

randomForest_rb <- caret::train(
  fantasyPoints_lag ~ ., # use all predictors
  data = seasonalData_lag_rb_train_imp$data,
  method = "rf",
  trControl = trainControl(
    method = "cv",
    number = 10)) # 10-fold cross-validation

randomForest_wr <- caret::train(
  fantasyPoints_lag ~ ., # use all predictors
  data = seasonalData_lag_wr_train_imp$data,
  method = "rf",
  trControl = trainControl(
    method = "cv",
    number = 10)) # 10-fold cross-validation

randomForest_te <- caret::train(
  fantasyPoints_lag ~ ., # use all predictors
  data = seasonalData_lag_te_train_imp$data,
  method = "rf",
  trControl = trainControl(
    method = "cv",
    number = 10)) # 10-fold cross-validation

stopCluster(cl)

print(randomForest_qb)
print(randomForest_rb)
print(randomForest_wr)
print(randomForest_te)

newData_qb$fantasyPoints_lag <- predict(
  randomForest_qb,
  newdata = newData_qb
)

newData_rb$fantasyPoints_lag <- predict(
  randomForest_rb,
  newdata = newData_rb
)

newData_wr$fantasyPoints_lag <- predict(
  randomForest_wr,
  newdata = newData_wr
)

newData_te$fantasyPoints_lag <- predict(
  randomForest_te,
  newdata = newData_te
)

newData_qb$fantasyPoints_lag <- newData_qb$fantasyPoints_lag
newData_rb$fantasyPoints_lag <- newData_rb$fantasyPoints_lag
newData_wr$fantasyPoints_lag <- newData_wr$fantasyPoints_lag
newData_te$fantasyPoints_lag <- newData_te$fantasyPoints_lag

newData_qb <- dplyr::left_join(
  newData_qb,
  newData_seasonal %>% select(gsis_id, player_display_name, team),
  by = "gsis_id"
)

newData_rb <- dplyr::left_join(
  newData_rb,
  newData_seasonal %>% select(gsis_id, player_display_name, team),
  by = "gsis_id"
)

newData_wr <- dplyr::left_join(
  newData_wr,
  newData_seasonal %>% select(gsis_id, player_display_name, team),
  by = "gsis_id"
)

newData_te <- dplyr::left_join(
  newData_te,
  newData_seasonal %>% select(gsis_id, player_display_name, team),
  by = "gsis_id"
)

newData_qb %>%
  arrange(-fantasyPoints_lag) %>% 
  select(gsis_id, player_display_name, fantasyPoints_lag)

newData_rb %>%
  arrange(-fantasyPoints_lag) %>% 
  select(gsis_id, player_display_name, fantasyPoints_lag)

newData_wr %>%
  arrange(-fantasyPoints_lag) %>% 
  select(gsis_id, player_display_name, fantasyPoints_lag)

newData_te %>%
  arrange(-fantasyPoints_lag) %>% 
  select(gsis_id, player_display_name, fantasyPoints_lag)
```

>>>>

```{r}
library("petersenlab")
library("parallel")
library("future")
library("missRanger")
library("powerjoin")
library("tidymodels")
library("LongituRF")
library("gpboost")
library("tidyverse")
```

@@

num_cores <- parallel::detectCores() - 1
num_true_cores <- parallel::detectCores(logical = FALSE) - 1
```

@@

We use the `future` [@R-future] package for parallel (faster) processing.

```{r}
future::plan(future::multisession, workers = num_cores)
```

@@

```{r}
# Set seed for reproducibility
set.seed(52242)

# Set up cross-validation
folds <- rsample::group_vfold_cv(
  data_train_qb,
  group = gsis_id,
  v = 10) # 10-fold cross-validation

# Define Recipe (Formula)
rec <- recipes::recipe(
  fantasyPoints_lag ~ ageCentered20,
  data = data_train_qb)

# Define Model
lm_spec <- parsnip::linear_reg() %>%
  parsnip::set_engine("lm") %>%
  parsnip::set_mode("regression")

# Workflow
lm_wf <- workflows::workflow() %>%
  workflows::add_recipe(rec) %>%
  workflows::add_model(lm_spec)

# Fit Model with Cross-Validation
cv_results <- tune::fit_resamples(
  lm_wf,
  resamples = folds,
  metrics = metric_set(rmse, mae, rsq),
  control = control_resamples(save_pred = TRUE)
)

# View Cross-Validation metrics
tune::collect_metrics(cv_results)

# Fit Final Model on Training Data
final_model <- workflows::fit(
  lm_wf,
  data = data_train_qb)

# View Coefficients
final_model %>% 
  workflows::extract_fit_parsnip() %>% 
  broom::tidy()

# Predict on Test Data
predict(final_model, data_test_qb)

df <- data_test_qb %>%
  mutate(pred = predict(final_model, new_data = data_test_qb)$.pred)

# Evaluate Accuracy of Predictions
petersenlab::accuracyOverall(
  predicted = df$pred,
  actual = df$fantasyPoints_lag,
  dropUndefined = TRUE
)
```

```{r}
#| label: fig-regressionOnePredictor
#| fig-cap: "Predicted Versus Actual Fantasy Points for Regression Model with One Predictor (Player Age)."
#| fig-alt: "Predicted Versus Actual Fantasy Points for Regression Model with One Predictor (Player Age)."

# Calculate combined range for axes
axis_limits <- range(c(df$pred, df$fantasyPoints_lag), na.rm = TRUE)

ggplot(
  df,
  aes(
    x = pred,
    y = fantasyPoints_lag)) +
  geom_point(
    size = 2,
    alpha = 0.6) +
  geom_abline(
    slope = 1,
    intercept = 0,
    color = "blue",
    linetype = "dashed") +
  coord_equal(
    xlim = axis_limits,
    ylim = axis_limits) +
  labs(
    title = "Predicted vs Actual Fantasy Points (Test Data)",
    x = "Predicted Fantasy Points",
    y = "Actual Fantasy Points"
  ) +
  theme_classic()
```

```{r}
newData_qb %>%
  mutate(fantasyPoints_lag = predict(final_model, new_data = newData_qb)$.pred) %>% 
  left_join(
    .,
    nfl_playerIDs %>% select(gsis_id, name),
    by = "gsis_id"
  ) %>% 
  select(name, fantasyPoints_lag) %>% 
  arrange(-fantasyPoints_lag)
```

@@

```{r}
# Set seed for reproducibility
set.seed(52242)

# Set up cross-validation
folds <- rsample::group_vfold_cv(
  data_train_qb,
  group = gsis_id,
  v = 10) # 10-fold cross-validation

# Define Recipe (Formula)
rec <- recipes::recipe(
  fantasyPoints_lag ~ .,
  data = data_train_qb %>% select(-gsis_id, -fantasyPointsMC_lag))

# Define Model
lm_spec <- parsnip::linear_reg() %>%
  parsnip::set_engine("lm") %>%
  parsnip::set_mode("regression")

# Workflow
lm_wf <- workflows::workflow() %>%
  workflows::add_recipe(rec) %>%
  workflows::add_model(lm_spec)

# Fit Model with Cross-Validation
cv_results <- tune::fit_resamples(
  lm_wf,
  resamples = folds,
  metrics = metric_set(rmse, mae, rsq),
  control = control_resamples(save_pred = TRUE)
)

# View Cross-Validation metrics
tune::collect_metrics(cv_results)

# Fit Final Model on Training Data
final_model <- workflows::fit(
  lm_wf,
  data = data_train_qb)

# View Coefficients
final_model %>% 
  workflows::extract_fit_parsnip() %>% 
  broom::tidy()

# Predict on Test Data
predict(final_model, data_test_qb)

df <- data_test_qb %>%
  mutate(pred = predict(final_model, new_data = data_test_qb)$.pred)

# Evaluate Accuracy of Predictions
petersenlab::accuracyOverall(
  predicted = df$pred,
  actual = df$fantasyPoints_lag,
  dropUndefined = TRUE
)
```

```{r}
#| label: fig-regressionMultiplePredictors
#| fig-cap: "Predicted Versus Actual Fantasy Points for Regression Model with Multiple Predictors."
#| fig-alt: "Predicted Versus Actual Fantasy Points for Regression Model with Multiple Predictors."

# Calculate combined range for axes
axis_limits <- range(c(df$pred, df$fantasyPoints_lag), na.rm = TRUE)

ggplot(
  df,
  aes(
    x = pred,
    y = fantasyPoints_lag)) +
  geom_point(
    size = 2,
    alpha = 0.6) +
  geom_abline(
    slope = 1,
    intercept = 0,
    color = "blue",
    linetype = "dashed") +
  coord_equal(
    xlim = axis_limits,
    ylim = axis_limits) +
  labs(
    title = "Predicted vs Actual Fantasy Points (Test Data)",
    x = "Predicted Fantasy Points",
    y = "Actual Fantasy Points"
  ) +
  theme_classic()
```

```{r}
newData_qb %>%
  mutate(fantasyPoints_lag = predict(final_model, new_data = newData_qb)$.pred) %>% 
  left_join(
    .,
    nfl_playerIDs %>% select(gsis_id, name),
    by = "gsis_id"
  ) %>% 
  select(name, fantasyPoints_lag) %>% 
  arrange(-fantasyPoints_lag)
```

@@

```{r}
# Set seed for reproducibility
set.seed(52242)

# Set up cross-validation
folds <- rsample::group_vfold_cv(
  data_train_qb,
  group = gsis_id,
  v = 10) # 10-fold cross-validation

# Define Recipe (Formula)
rec <- recipes::recipe(
  fantasyPoints_lag ~ .,
  data = data_train_qb %>% select(-gsis_id, -fantasyPointsMC_lag))

# Define Model
lasso_spec <- 
  parsnip::linear_reg(
    penalty = tune(),
    mixture = 1) %>%
  set_engine("glmnet")

# Workflow
lasso_wf <- workflows::workflow() %>%
  workflows::add_recipe(rec) %>%
  workflows::add_model(lasso_spec)

# Define grid of penalties to try (log scale is typical)
penalty_grid <- dials::grid_regular(
  dials::penalty(range = c(-4, -1)),
  levels = 20)

# Tune the Penalty Parameter
cv_results <- tune::tune_grid(
  lasso_wf,
  resamples = folds,
  grid = penalty_grid,
  metrics = metric_set(rmse, mae, rsq),
  control = control_grid(save_pred = TRUE)
)

# View Cross-Validation metrics
tune::collect_metrics(cv_results)

# Identify best penalty
select_best(cv_results, metric = "rmse")
select_best(cv_results, metric = "mae")
select_best(cv_results, metric = "rsq")

best_penalty <- select_best(cv_results, metric = "mae")

# Finalize Workflow with Best Penalty
final_wf <- finalize_workflow(
  lasso_wf,
  best_penalty)

# Fit Final Model on Training Data
final_model <- workflows::fit(
  final_wf,
  data = data_train_qb)

# View Coefficients
final_model %>% 
  workflows::extract_fit_parsnip() %>% 
  broom::tidy()

# Predict on Test Data
predict(final_model, data_test_qb)

df <- data_test_qb %>%
  mutate(pred = predict(final_model, new_data = data_test_qb)$.pred)

# Evaluate Accuracy of Predictions
petersenlab::accuracyOverall(
  predicted = df$pred,
  actual = df$fantasyPoints_lag,
  dropUndefined = TRUE
)
```

```{r}
#| label: fig-lassoModel
#| fig-cap: "Predicted Versus Actual Fantasy Points for Least Absolute Shrinkage and Selection Option (LASSO) Model."
#| fig-alt: "Predicted Versus Actual Fantasy Points for Least Absolute Shrinkage and Selection Option (LASSO) Model."

# Calculate combined range for axes
axis_limits <- range(c(df$pred, df$fantasyPoints_lag), na.rm = TRUE)

ggplot(
  df,
  aes(
    x = pred,
    y = fantasyPoints_lag)) +
  geom_point(
    size = 2,
    alpha = 0.6) +
  geom_abline(
    slope = 1,
    intercept = 0,
    color = "blue",
    linetype = "dashed") +
  coord_equal(
    xlim = axis_limits,
    ylim = axis_limits) +
  labs(
    title = "Predicted vs Actual Fantasy Points (Test Data)",
    x = "Predicted Fantasy Points",
    y = "Actual Fantasy Points"
  ) +
  theme_classic()
```

```{r}
newData_qb %>%
  mutate(fantasyPoints_lag = predict(final_model, new_data = newData_qb)$.pred) %>% 
  left_join(
    .,
    nfl_playerIDs %>% select(gsis_id, name),
    by = "gsis_id"
  ) %>% 
  select(name, fantasyPoints_lag) %>% 
  arrange(-fantasyPoints_lag)
```

@@

```{r}
# Set seed for reproducibility
set.seed(52242)

# Set up cross-validation
folds <- rsample::group_vfold_cv(
  data_train_qb,
  group = gsis_id,
  v = 10) # 10-fold cross-validation

# Define Recipe (Formula)
rec <- recipes::recipe(
  fantasyPoints_lag ~ .,
  data = data_train_qb %>% select(-gsis_id, -fantasyPointsMC_lag))

# Define Model
ridge_spec <- 
  linear_reg(
    penalty = tune(),
    mixture = 0) %>%
  set_engine("glmnet")

# Workflow
ridge_wf <- workflows::workflow() %>%
  workflows::add_recipe(rec) %>%
  workflows::add_model(ridge_spec)

# Define grid of penalties to try (log scale is typical)
penalty_grid <- dials::grid_regular(
  dials::penalty(range = c(-4, -1)),
  levels = 20)

# Tune the Penalty Parameter
cv_results <- tune::tune_grid(
  ridge_wf,
  resamples = folds,
  grid = penalty_grid,
  metrics = metric_set(rmse, mae, rsq),
  control = control_grid(save_pred = TRUE)
)

# View Cross-Validation metrics
tune::collect_metrics(cv_results)

# Identify best penalty
select_best(cv_results, metric = "rmse")
select_best(cv_results, metric = "mae")
select_best(cv_results, metric = "rsq")

best_penalty <- select_best(cv_results, metric = "mae")

# Finalize Workflow with Best Penalty
final_wf <- finalize_workflow(
  ridge_wf,
  best_penalty)

# Fit Final Model on Training Data
final_model <- workflows::fit(
  final_wf,
  data = data_train_qb)

# View Coefficients
final_model %>% 
  workflows::extract_fit_parsnip() %>% 
  broom::tidy()

# Predict on Test Data
predict(final_model, data_test_qb)

df <- data_test_qb %>%
  mutate(pred = predict(final_model, new_data = data_test_qb)$.pred)

# Evaluate Accuracy of Predictions
petersenlab::accuracyOverall(
  predicted = df$pred,
  actual = df$fantasyPoints_lag,
  dropUndefined = TRUE
)
```

```{r}
#| label: fig-ridgeRegressionModel
#| fig-cap: "Predicted Versus Actual Fantasy Points for Ridge Regression Model."
#| fig-alt: "Predicted Versus Actual Fantasy Points for Ridge Regression Model."

# Calculate combined range for axes
axis_limits <- range(c(df$pred, df$fantasyPoints_lag), na.rm = TRUE)

ggplot(
  df,
  aes(
    x = pred,
    y = fantasyPoints_lag)) +
  geom_point(
    size = 2,
    alpha = 0.6) +
  geom_abline(
    slope = 1,
    intercept = 0,
    color = "blue",
    linetype = "dashed") +
  coord_equal(
    xlim = axis_limits,
    ylim = axis_limits) +
  labs(
    title = "Predicted vs Actual Fantasy Points (Test Data)",
    x = "Predicted Fantasy Points",
    y = "Actual Fantasy Points"
  ) +
  theme_classic()
```

```{r}
newData_qb %>%
  mutate(fantasyPoints_lag = predict(final_model, new_data = newData_qb)$.pred) %>% 
  left_join(
    .,
    nfl_playerIDs %>% select(gsis_id, name),
    by = "gsis_id"
  ) %>% 
  select(name, fantasyPoints_lag) %>% 
  arrange(-fantasyPoints_lag)
```

@@

```{r}
# Set seed for reproducibility
set.seed(52242)

# Set up cross-validation
folds <- rsample::group_vfold_cv(
  data_train_qb,
  group = gsis_id,
  v = 10) # 10-fold cross-validation

# Define Recipe (Formula)
rec <- recipes::recipe(
  fantasyPoints_lag ~ .,
  data = data_train_qb %>% select(-gsis_id, -fantasyPointsMC_lag))

# Define Model
enet_spec <- 
  linear_reg(
    penalty = tune(),
    mixture = tune()) %>%
  set_engine("glmnet")

# Workflow
enet_wf <- workflows::workflow() %>%
  workflows::add_recipe(rec) %>%
  workflows::add_model(enet_spec)

# Define a regular grid for both penalty and mixture
grid_enet <- dials::grid_regular(
  dials::penalty(range = c(-4, -1)),
  dials::mixture(range = c(0, 1)),
  levels = c(20, 5) # 20 penalty values × 5 mixture values
)

# Tune the Grid
cv_results <- tune::tune_grid(
  enet_wf,
  resamples = folds,
  grid = grid_enet,
  metrics = metric_set(rmse, mae, rsq),
  control = control_grid(save_pred = TRUE)
)

# View Cross-Validation metrics
tune::collect_metrics(cv_results)

# Identify best penalty
select_best(cv_results, metric = "rmse")
select_best(cv_results, metric = "mae")
select_best(cv_results, metric = "rsq")

best_penalty <- select_best(cv_results, metric = "mae")

# Finalize Workflow with Best Penalty
final_wf <- finalize_workflow(
  enet_wf,
  best_penalty)

# Fit Final Model on Training Data
final_model <- workflows::fit(
  final_wf,
  data = data_train_qb)

# View Coefficients
final_model %>% 
  workflows::extract_fit_parsnip() %>% 
  broom::tidy()

# Predict on Test Data
predict(final_model, data_test_qb)

df <- data_test_qb %>%
  mutate(pred = predict(final_model, new_data = data_test_qb)$.pred)

# Evaluate Accuracy of Predictions
petersenlab::accuracyOverall(
  predicted = df$pred,
  actual = df$fantasyPoints_lag,
  dropUndefined = TRUE
)
```

```{r}
#| label: fig-elasticNetModel
#| fig-cap: "Predicted Versus Actual Fantasy Points for Elastic Net Model."
#| fig-alt: "Predicted Versus Actual Fantasy Points for Elastic Net Model."

# Calculate combined range for axes
axis_limits <- range(c(df$pred, df$fantasyPoints_lag), na.rm = TRUE)

ggplot(
  df,
  aes(
    x = pred,
    y = fantasyPoints_lag)) +
  geom_point(
    size = 2,
    alpha = 0.6) +
  geom_abline(
    slope = 1,
    intercept = 0,
    color = "blue",
    linetype = "dashed") +
  coord_equal(
    xlim = axis_limits,
    ylim = axis_limits) +
  labs(
    title = "Predicted vs Actual Fantasy Points (Test Data)",
    x = "Predicted Fantasy Points",
    y = "Actual Fantasy Points"
  ) +
  theme_classic()
```

```{r}
newData_qb %>%
  mutate(fantasyPoints_lag = predict(final_model, new_data = newData_qb)$.pred) %>% 
  left_join(
    .,
    nfl_playerIDs %>% select(gsis_id, name),
    by = "gsis_id"
  ) %>% 
  select(name, fantasyPoints_lag) %>% 
  arrange(-fantasyPoints_lag)
```

@@

```{r}
# Set seed for reproducibility
set.seed(52242)

# Set up cross-validation
folds <- rsample::group_vfold_cv(
  data_train_qb,
  group = gsis_id,
  v = 10) # 10-fold cross-validation

# Define Recipe (Formula)
rec <- recipes::recipe(
  fantasyPoints_lag ~ .,
  data = data_train_qb %>% select(-gsis_id, -fantasyPointsMC_lag))

# Define Model
rf_spec <- 
  parsnip::rand_forest(
    mtry = tune::tune(),
    min_n = tune::tune(),
    trees = 500) %>%
  parsnip::set_mode("regression") %>%
  parsnip::set_engine("ranger", importance = "impurity")

# Workflow
rf_wf <- workflows::workflow() %>%
  workflows::add_recipe(rec) %>%
  workflows::add_model(rf_spec)

# Create Grid
n_predictors <- recipes::prep(rec) %>%
  recipes::juice() %>%
  dplyr::select(-fantasyPoints_lag) %>%
  ncol()

# Dynamically define ranges based on data
rf_params <- hardhat::extract_parameter_set_dials(rf_spec) %>%
  dials:::update.parameters(
    mtry = dials::mtry(range = c(1L, n_predictors)),
    min_n = dials::min_n(range = c(2L, 10L))
  )

rf_grid <- dials::grid_random(rf_params, size = 15) #dials::grid_regular(rf_params, levels = 5)

# Tune the Grid
cv_results <- tune::tune_grid(
  rf_wf,
  resamples = folds,
  grid = rf_grid,
  metrics = metric_set(rmse, mae, rsq),
  control = control_grid(save_pred = TRUE)
)

# View Cross-Validation metrics
tune::collect_metrics(cv_results)

# Identify best penalty
select_best(cv_results, metric = "rmse")
select_best(cv_results, metric = "mae")
select_best(cv_results, metric = "rsq")

best_penalty <- select_best(cv_results, metric = "mae")

# Finalize Workflow with Best Penalty
final_wf <- finalize_workflow(
  rf_wf,
  best_penalty)

# Fit Final Model on Training Data
final_model <- workflows::fit(
  final_wf,
  data = data_train_qb)

# View Feature Importance
rf_fit <- final_model %>% 
  workflows::extract_fit_parsnip()

rf_fit

ranger_obj <- rf_fit$fit

ranger_obj$variable.importance

# Predict on Test Data
predict(final_model, data_test_qb)

df <- data_test_qb %>%
  mutate(pred = predict(final_model, new_data = data_test_qb)$.pred)

# Evaluate Accuracy of Predictions
petersenlab::accuracyOverall(
  predicted = df$pred,
  actual = df$fantasyPoints_lag,
  dropUndefined = TRUE
)
```

```{r}
#| label: fig-randomForestModel
#| fig-cap: "Predicted Versus Actual Fantasy Points for Random Forest Model."
#| fig-alt: "Predicted Versus Actual Fantasy Points for Random Forest Model."

# Calculate combined range for axes
axis_limits <- range(c(df$pred, df$fantasyPoints_lag), na.rm = TRUE)

ggplot(
  df,
  aes(
    x = pred,
    y = fantasyPoints_lag)) +
  geom_point(
    size = 2,
    alpha = 0.6) +
  geom_abline(
    slope = 1,
    intercept = 0,
    color = "blue",
    linetype = "dashed") +
  coord_equal(
    xlim = axis_limits,
    ylim = axis_limits) +
  labs(
    title = "Predicted vs Actual Fantasy Points (Test Data)",
    x = "Predicted Fantasy Points",
    y = "Actual Fantasy Points"
  ) +
  theme_classic()
```

```{r}
newData_qb %>%
  mutate(fantasyPoints_lag = predict(final_model, new_data = newData_qb)$.pred) %>% 
  left_join(
    .,
    nfl_playerIDs %>% select(gsis_id, name),
    by = "gsis_id"
  ) %>% 
  select(name, fantasyPoints_lag) %>% 
  arrange(-fantasyPoints_lag)
```

Now we can stop the parallel backend:

```{r}
future::plan(future::sequential)
```

```{r}
#| include: false
